---
title: "Delivery & Newborn Claims Counts vs. TX Vital Statistics"
format: html
---

## Overview

Compares two independent claims-based counts against `colour.ref_vital_stats_tx`
(CDC WONDER TX birth counts by source of payment for delivery), each of which
should approximate vital-stats **births** in its own way:

- **Deliveries** (`colour.delivery_records_insurance_stg`, built from the
  `Delivery` tab of the Orr 2024 codelist, age > 12) - a person-year-grain
  staging table. Two counts are pulled from it, since a delivery episode and
  a birth aren't the same thing once twins are in the mix:
  - **`n_mothers`** - `COUNT(DISTINCT pers_id)`, i.e. distinct people with a
    qualifying delivery in that year
  - **`n_babies`** - the same deliveries, summed by `baby_count`, so a set of
    twins contributes 2. This is the count that should line up with vital
    stats births, while `n_mothers` will run slightly below it.
- **Newborns** (`colour.newborn_records_insurance_stg`, built from the
  `newborn` tab of the codelist, age < 1) - a person-year-grain staging
  table. Only one count applies here:
  - **`n_newborns`** - `COUNT(DISTINCT pers_id)`, i.e. distinct newborns
    identified in claims that year

Both pipelines assign insurance the same way - `prim_med_plan` from
`research_di.agg_yr_plan` for that year (an enrollment-based check) - so the
`ins` categories line up 1:1 across deliveries, newborns, and vital stats:

| Claims (`ins`) | `prim_med_plan` values                                                              | Vital stats                                  |
|-----------------|--------------------------------------------------------------------------------------|-----------------------------------------------|
| Medicaid         | Medicaid                                                                              | `medicaid`                                     |
| Commercial       | Commercial, Com-Ers, Com-ErsTrs, Com-Trs                                              | `private_insurance`                            |
| Other            | Medicare Advantage(+ Imputed), Medicare FFS, Medicare Imputed, Federal, Not Enrolled  | `self_pay` + `other` + `unknown_or_not_stated` |
| Total            | (all of the above)                                                                    | `total`                                        |

Both tables below carry a per-year **Total** row (summed across insurance
categories) alongside the by-category breakdown, and both report a raw
difference (`claims - vital`) and a % difference (`diff / vital`) so
over/under-capture is easy to spot at a glance.

```{r}
#| label: setup
#| message: false

pacman::p_load(RPostgres, DBI, keyring, glue, tidyverse, scales, gt, openxlsx)
```

```{r}
#| label: connect

greenplum_user <- "jwozny"

uthealth <- dbConnect(
  RPostgres::Postgres(),
  dbname   = "apcd_dev",
  user     = greenplum_user,
  password = keyring::key_get("apcd_dev", greenplum_user),
  host     = "spgpcdplv001",
  bigint   = "integer"
)
```

```{r}
#| label: pull-vital-stats

vital_stats <- dbGetQuery(uthealth, "
  SELECT * FROM colour.ref_vital_stats_tx
  WHERE year >= 2019
  ORDER BY year
") |>
  as_tibble()

vital_stats
```

```{r}
#| label: reshape-vital-stats

# Collapse to the same 4 buckets (3 insurance categories + Total) that both
# claims-side tables use, so joins below are a straight (year, ins) match.
vital_long <- vital_stats |>
  transmute(
    year,
    Medicaid   = medicaid,
    Commercial = private_insurance,
    Other      = self_pay + other + unknown_or_not_stated,
    Total      = total
  ) |>
  pivot_longer(
    cols = c(Medicaid, Commercial, Other, Total),
    names_to  = "ins",
    values_to = "vital_n"
  )

vital_long
```

```{r}
#| label: ins-order

# Shared factor levels so both output tables sort Medicaid / Commercial /
# Other / Total consistently within each year.
ins_levels <- c("Medicaid", "Commercial", "Other", "Total")
```

## Deliveries (Mothers + Babies)

```{r}
#| label: pull-delivery-counts

delivery_claims <- dbGetQuery(uthealth, "
  SELECT year, ins,
         count(DISTINCT pers_id) AS n_mothers,
         sum(baby_count)         AS n_babies
  FROM colour.delivery_records_insurance_stg
  GROUP BY year, ins
  ORDER BY year, ins
") |>
  as_tibble() |>
  # Fold the QC-only 'Not Enrolled' bucket into 'Other' so the categories
  # line up 1:1 with the 3-way vital stats split.
  mutate(ins = if_else(ins == "Not Enrolled", "Other", ins)) |>
  group_by(year, ins) |>
  summarise(
    n_mothers = sum(n_mothers),
    n_babies  = sum(n_babies),
    .groups = "drop"
  )

delivery_claims
```

```{r}
#| label: delivery-add-total-row

delivery_total <- delivery_claims |>
  group_by(year) |>
  summarise(
    n_mothers = sum(n_mothers),
    n_babies  = sum(n_babies),
    .groups = "drop"
  ) |>
  mutate(ins = "Total")

delivery_all <- bind_rows(delivery_claims, delivery_total) |>
  mutate(ins = factor(ins, levels = ins_levels)) |>
  arrange(year, ins)

delivery_all
```

```{r}
#| label: delivery-compare

delivery_comparison <- delivery_all |>
  left_join(vital_long, by = c("year", "ins")) |>
  mutate(
    diff_mothers     = n_mothers - vital_n,
    pct_diff_mothers = diff_mothers / vital_n,
    diff_babies      = n_babies - vital_n,
    pct_diff_babies  = diff_babies / vital_n
  ) |>
  arrange(year, ins)

delivery_comparison
```

```{r}
#| label: table-deliveries

delivery_comparison |>
  mutate(
    pct_diff_mothers = percent(pct_diff_mothers, accuracy = 0.1),
    pct_diff_babies  = percent(pct_diff_babies, accuracy = 0.1)
  ) |>
  gt(groupname_col = "year") |>
  tab_header(title = "Claims-Identified Deliveries vs. TX Vital Statistics Births") |>
  cols_label(
    ins               = "Insurance Category",
    n_mothers         = "Claims Mothers",
    n_babies          = "Claims Births (baby_count-adj.)",
    vital_n           = "Vital Stats Births",
    diff_mothers      = "Diff (Mothers)",
    pct_diff_mothers  = "% Diff (Mothers)",
    diff_babies       = "Diff (Births)",
    pct_diff_babies   = "% Diff (Births)"
  ) |>
  fmt_number(columns = c(n_mothers, n_babies, vital_n, diff_mothers, diff_babies), decimals = 0)
```

```{r}
#| label: fig-deliveries
#| fig-cap: "Claims-identified births (baby_count-adjusted) vs. vital statistics births, by insurance category"

delivery_comparison |>
  filter(ins != "Total") |>
  pivot_longer(cols = c(n_babies, vital_n), names_to = "source", values_to = "n") |>
  mutate(source = recode(source, n_babies = "Claims (Births)", vital_n = "Vital Stats")) |>
  ggplot(aes(x = year, y = n, color = source)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ ins, scales = "free_y") +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Count", color = NULL) +
  theme_minimal()
```

## Newborns

```{r}
#| label: pull-newborn-counts

newborn_claims <- dbGetQuery(uthealth, "
  SELECT year, ins,
         count(DISTINCT pers_id) AS n_newborns
  FROM colour.newborn_records_insurance_stg
  GROUP BY year, ins
  ORDER BY year, ins
") |>
  as_tibble() |>
  mutate(ins = if_else(ins == "Not Enrolled", "Other", ins)) |>
  group_by(year, ins) |>
  summarise(
    n_newborns = sum(n_newborns),
    .groups = "drop"
  )

newborn_claims
```

```{r}
#| label: newborn-add-total-row

newborn_total <- newborn_claims |>
  group_by(year) |>
  summarise(
    n_newborns = sum(n_newborns),
    .groups = "drop"
  ) |>
  mutate(ins = "Total")

newborn_all <- bind_rows(newborn_claims, newborn_total) |>
  mutate(ins = factor(ins, levels = ins_levels)) |>
  arrange(year, ins)

newborn_all
```

```{r}
#| label: newborn-compare

newborn_comparison <- newborn_all |>
  left_join(vital_long, by = c("year", "ins")) |>
  mutate(
    diff_newborns     = n_newborns - vital_n,
    pct_diff_newborns = diff_newborns / vital_n
  ) |>
  arrange(year, ins)

newborn_comparison
```

```{r}
#| label: table-newborns

newborn_comparison |>
  mutate(pct_diff_newborns = percent(pct_diff_newborns, accuracy = 0.1)) |>
  gt(groupname_col = "year") |>
  tab_header(title = "Claims-Identified Newborns vs. TX Vital Statistics Births") |>
  cols_label(
    ins               = "Insurance Category",
    n_newborns        = "Claims Newborns",
    vital_n           = "Vital Stats Births",
    diff_newborns     = "Diff (Newborns)",
    pct_diff_newborns = "% Diff (Newborns)"
  ) |>
  fmt_number(columns = c(n_newborns, vital_n, diff_newborns), decimals = 0)
```

```{r}
#| label: fig-newborns
#| fig-cap: "Claims-identified newborns vs. vital statistics births, by insurance category"

newborn_comparison |>
  filter(ins != "Total") |>
  pivot_longer(cols = c(n_newborns, vital_n), names_to = "source", values_to = "n") |>
  mutate(source = recode(source, n_newborns = "Claims (Newborns)", vital_n = "Vital Stats")) |>
  ggplot(aes(x = year, y = n, color = source)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ ins, scales = "free_y") +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Count", color = NULL) +
  theme_minimal()
```

## Overall Summary

Combines just the per-year **Total** rows from both pipelines (all insurance
categories combined) into one high-level table.

```{r}
#| label: overall-compare

overall_comparison <- delivery_comparison |>
  filter(ins == "Total") |>
  select(year, vital_n, n_mothers, diff_mothers, pct_diff_mothers,
         n_babies, diff_babies, pct_diff_babies) |>
  left_join(
    newborn_comparison |>
      filter(ins == "Total") |>
      select(year, n_newborns, diff_newborns, pct_diff_newborns),
    by = "year"
  )

overall_comparison
```

```{r}
#| label: table-overall

overall_comparison |>
  mutate(
    pct_diff_mothers  = percent(pct_diff_mothers, accuracy = 0.1),
    pct_diff_babies   = percent(pct_diff_babies, accuracy = 0.1),
    pct_diff_newborns = percent(pct_diff_newborns, accuracy = 0.1)
  ) |>
  gt() |>
  tab_header(title = "Overall (All Insurance Categories Combined) vs. TX Vital Statistics Births") |>
  cols_label(
    year              = "Year",
    vital_n           = "Vital Stats Births",
    n_mothers         = "Claims Mothers",
    diff_mothers      = "Diff (Mothers)",
    pct_diff_mothers  = "% Diff (Mothers)",
    n_babies          = "Claims Births (baby_count-adj.)",
    diff_babies       = "Diff (Births)",
    pct_diff_babies   = "% Diff (Births)",
    n_newborns        = "Claims Newborns",
    diff_newborns     = "Diff (Newborns)",
    pct_diff_newborns = "% Diff (Newborns)"
  ) |>
  fmt_number(
    columns = c(vital_n, n_mothers, diff_mothers, n_babies, diff_babies, n_newborns, diff_newborns),
    decimals = 0
  )
```

## Export to Excel

Four tabs, each a real Excel Table (banded rows + filter dropdowns, via
`writeDataTable()`) except Steps, which is documentation:

1. **Overall** - Total-row-only summary across both pipelines (first tab)
2. **Deliveries** - mother count + baby count vs. vital stats, by category + Total
3. **Newborns** - person count vs. vital stats, by category + Total
4. **Steps** - methodology notes (last tab)

```{r}
#| label: export-to-excel

# -- EDIT THIS --------------------------------------------------
out_path <- here::here("claims_vs_vital_stats_comparison.xlsx")
# -----------------------------------------------------------------

wb <- createWorkbook()

table_style <- "TableStyleMedium9"

addWorksheet(wb, "Overall")
writeDataTable(wb, "Overall", overall_comparison, tableStyle = table_style)

addWorksheet(wb, "Deliveries")
writeDataTable(wb, "Deliveries", delivery_comparison, tableStyle = table_style)

addWorksheet(wb, "Newborns")
writeDataTable(wb, "Newborns", newborn_comparison, tableStyle = table_style)

# writeDataTable doesn't take a numFmt arg, so percent formatting is
# applied as a separate style pass on top of the table
pct_style <- createStyle(numFmt = "0.0%")

addStyle(
  wb, "Overall", pct_style,
  rows = 2:(nrow(overall_comparison) + 1),
  cols = which(names(overall_comparison) %in% c("pct_diff_mothers", "pct_diff_babies", "pct_diff_newborns")),
  gridExpand = TRUE
)
addStyle(
  wb, "Deliveries", pct_style,
  rows = 2:(nrow(delivery_comparison) + 1),
  cols = which(names(delivery_comparison) %in% c("pct_diff_mothers", "pct_diff_babies")),
  gridExpand = TRUE
)
addStyle(
  wb, "Newborns", pct_style,
  rows = 2:(nrow(newborn_comparison) + 1),
  cols = which(names(newborn_comparison) %in% c("pct_diff_newborns")),
  gridExpand = TRUE
)

# openxlsx's widths = "auto" is unreliable about leaving room for the
# header text itself (it can underestimate and clip column names), so
# size each column explicitly off the wider of the header name or its
# longest formatted value instead.
set_widths_for_headers <- function(wb, sheet, df) {
  col_widths <- vapply(seq_along(df), function(i) {
    header_len <- nchar(names(df)[i])
    data_len   <- max(nchar(format(df[[i]])), na.rm = TRUE)
    max(header_len, data_len) + 2
  }, numeric(1))
  setColWidths(wb, sheet, cols = seq_along(df), widths = col_widths)
}

set_widths_for_headers(wb, "Overall", overall_comparison)
set_widths_for_headers(wb, "Deliveries", delivery_comparison)
set_widths_for_headers(wb, "Newborns", newborn_comparison)

# ------------------------------------------------------------------
# Steps tab (methodology notes - plain text, not a data table)
# ------------------------------------------------------------------

steps_text <- c(
  "Get counts and compare to vital stats",
  "",
  "Infants:",
  "- Get all claims using \"Newborn Codes\" tab of Orr codelists (CPT and DX)",
  "- Filter ids to age < 1 in enrollment table",
  "- Get count distinct people",
  "",
  "Deliveries:",
  "- Get all claims using \"Delivery Codes\" tab of Orr codelists (CPT and DX)",
  "- Filter ids to those age > 12 in enrollment table (or left join to get those whose info is missing for the year)",
  "- Get one instance per person per year taking the maximum baby count for that year (for non-babycount codes, coalesce value of 1)",
  "",
  "Notes:",
  "- There is overlap of codes between deliveries and codesets",
  "- Vital Stats are recorded as the mother's insurance (not necessarily the baby's)",
  "- NY Vital Stats: Link",
  "- TX Vital Stats:"
)
header_rows <- c(1, 3, 8, 13)
tx_row      <- length(steps_text)

# -- EDIT THIS: the jsessionid pasted in chat is a temporary session
# token that will have expired - swap in a fresh/permanent query URL
# if you have one -----------------------------------------------
tx_url <- "https://wonder.cdc.gov/controller/datarequest/D149"
# -----------------------------------------------------------------

addWorksheet(wb, "Steps")
writeData(wb, "Steps", steps_text, colNames = FALSE)
addStyle(wb, "Steps", createStyle(textDecoration = "bold", fontSize = 12), rows = header_rows, cols = 1)

class(tx_url) <- "hyperlink"
writeData(wb, "Steps", x = tx_url, startCol = 2, startRow = tx_row)

setColWidths(wb, "Steps", cols = 1, widths = 100)
setColWidths(wb, "Steps", cols = 2, widths = 55)

saveWorkbook(wb, out_path, overwrite = TRUE)

cat("Wrote comparison workbook to", out_path, "\n")
```

```{r}
#| label: disconnect

dbDisconnect(uthealth)
```
