---
title: "Infant and Delivery Identification: Enrollment/Claim/Delivery Windows"
format:
  docx:
    toc: false
    toc-depth: 2
    fig-width: 7
    fig-height: 4.5
execute:
  echo: false
  warning: false
  message: false
---

## Overview

Infants start from the DOB spine: every person with a single DOB
on or after 2019-01-01. Newborn codes are measured against that spine rather
than defining it. Two requirements apply in sequence:

1. Enrollment begins within 1-12 months of the DOB.
2. A newborn-code claim appears within 7, 30, 60 or 90 days of the DOB, among
   infants who cleared step 1.

Deliveries are pseudo-episodes: claims collapsed on overlapping date
spans, anchored on the CPT procedure date where one exists. 

```{r}
#| label: setup

pacman::p_load(RPostgres, DBI, keyring, glue, tidyverse, scales,
               flextable, officer, openxlsx, here)

# gt's Word output is unreliable under Quarto docx, so flextable.
set_flextable_defaults(font.size = 9, padding = 3)

ft <- function(df, caption = NULL) {
  x <- flextable(df) |>
    theme_booktabs() |>
    align(align = "right", part = "all") |>
    align(j = 1, align = "left", part = "all") |>
    autofit()
  if (!is.null(caption)) x <- set_caption(x, caption)
  x
}
```

```{r}
#| label: connect

greenplum_user <- "jwozny"

uthealth <- dbConnect(
  RPostgres::Postgres(),
  dbname   = "apcd_dev",
  user     = greenplum_user,
  password = keyring::key_get("apcd_dev", greenplum_user),
  host     = "spgpcdplv001",
  bigint   = "numeric"
)
```

```{r}
#| label: parameters

# Windows used by the headline tables and the Excel export. The sweeps
# always cover the full grid regardless.
enroll_window_months <- 12    # 1-12
nb_claim_window_days <- 30    # 7, 30, 60, 90

# A window month is 30 days, so 12 months = 360. SUMMARY B in
# create_newborn_records_v2.sql uses 365.

year_min <- 2019
year_max <- 2025

# Last birth year with 12 months of follow-up. Later cohorts are censored,
# not low, so the figures drop them.
last_complete_birth_year <- 2024

# Shortest plausible interval between separate deliveries.
min_preg_gap_days <- 270
# -----------------------------------------------------------------

enroll_window_days <- enroll_window_months * 30
ins_levels <- c("Medicaid", "Commercial", "Other", "Total")
complete_years_label <- glue("{year_min}-{last_complete_birth_year}")
```

```{r}
#| label: helpers

# 'Not Enrolled' folded into 'Other' to match the 3-way vital stats split.
fold_not_enrolled <- function(df, keys, value_cols) {
  df |>
    mutate(ins = if_else(ins == "Not Enrolled", "Other", ins)) |>
    group_by(across(all_of(c(keys, "ins")))) |>
    summarise(across(all_of(value_cols), sum), .groups = "drop")
}

add_ins_total <- function(df, keys, value_cols) {
  totals <- df |>
    group_by(across(all_of(keys))) |>
    summarise(across(all_of(value_cols), sum), .groups = "drop") |>
    mutate(ins = "Total")

  bind_rows(df, totals) |>
    mutate(ins = factor(ins, levels = ins_levels)) |>
    arrange(!!!syms(c(keys, "ins")))
}

ins_colours <- c(Medicaid = "#1b7837", Commercial = "#2166ac",
                 Other = "#b2182b", Total = "grey30")
scale_ins <- function() scale_colour_manual(values = ins_colours, drop = FALSE)
```

# Infants

## Requirements

Both windows applied in sequence at the headline settings. 

```{r}
#| label: pull-waterfall

waterfall <- dbGetQuery(uthealth, glue("
  SELECT birth_year,
         COUNT(*) AS n_spine,
         SUM(CASE WHEN days_to_enroll IS NOT NULL THEN 1 ELSE 0 END) AS n_ever_enrolled,
         SUM(CASE WHEN days_to_enroll <= {enroll_window_days} THEN 1 ELSE 0 END) AS n_enrolled_in_window,
         SUM(CASE WHEN days_to_enroll <= {enroll_window_days}
                   AND days_to_first_nb_claim <= {nb_claim_window_days}
                  THEN 1 ELSE 0 END) AS n_with_nb_claim
  FROM colour.infant_records
  WHERE birth_year BETWEEN {year_min} AND {year_max}
  GROUP BY 1 ORDER BY 1
")) |> as_tibble()

waterfall |>
  mutate(
    pct_enrolled = n_enrolled_in_window / n_spine,
    pct_claim    = n_with_nb_claim / n_enrolled_in_window
  ) |>
  mutate(across(starts_with("n_"), ~ comma(.x, accuracy = 1)),
         across(starts_with("pct_"), ~ percent(.x, accuracy = 0.1))) |>
  ft(glue("Infant waterfall at {enroll_window_months}-month enrollment and ",
          "{nb_claim_window_days}-day claim windows. n_ever_enrolled ignores ",
          "timing; n_enrolled_in_window applies it. pct_claim is conditional on ",
          "clearing the enrollment step."))
```

## Enrollment window (out of everyone with a DOB)

Insurance is the plan on the segment closest to the DOB. It exists for anyone
who ever enrolls, whenever that happens, so a late enroller still has a category
and the window filter does not determine it.

```{r}
#| label: pull-enroll-sweep

enroll_sweep_raw <- dbGetQuery(uthealth, glue("
  SELECT m.months,
         r.birth_year,
         r.ins,
         COUNT(*) AS n_in_group,
         SUM(CASE WHEN r.days_to_enroll <= m.months * 30 THEN 1 ELSE 0 END) AS n_enrolled
  FROM colour.infant_records r
  CROSS JOIN generate_series(1, 12) AS m(months)
  WHERE r.birth_year BETWEEN {year_min} AND {year_max}
  GROUP BY 1, 2, 3
  ORDER BY 1, 2, 3
")) |> as_tibble()

spine_by_year <- enroll_sweep_raw |>
  filter(months == 1) |>
  group_by(birth_year) |>
  summarise(n_spine = sum(n_in_group), .groups = "drop")

spine_complete <- spine_by_year |>
  filter(birth_year <= last_complete_birth_year) |>
  summarise(n = sum(n_spine)) |>
  pull(n)

# 'Not Enrolled' is the never-enrolled bucket, i.e. the complement of these
# curves. It sits in the waterfall instead.
enroll_sweep <- enroll_sweep_raw |>
  filter(ins != "Not Enrolled") |>
  add_ins_total(keys = c("months", "birth_year"),
                value_cols = c("n_in_group", "n_enrolled")) |>
  left_join(spine_by_year, by = "birth_year") |>
  group_by(birth_year, ins) |>
  mutate(pct_of_group_final = n_enrolled / n_enrolled[months == 12]) |>
  ungroup() |>
  mutate(pct_of_spine = n_enrolled / n_spine)

# Pooled across complete cohorts, for the figures.
enroll_pooled <- enroll_sweep |>
  filter(birth_year <= last_complete_birth_year) |>
  group_by(months, ins) |>
  summarise(n_enrolled = sum(n_enrolled), .groups = "drop") |>
  group_by(ins) |>
  mutate(pct_of_group_final = n_enrolled / n_enrolled[months == 12]) |>
  ungroup() |>
  mutate(pct_of_spine = n_enrolled / spine_complete)
```

Share of the birth cohort enrolled, by window.

```{r}
#| label: fig-enroll-share

enroll_pooled |>
  ggplot(aes(x = months, y = pct_of_spine, colour = ins)) +
  geom_line() +
  geom_point(size = 1) +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(labels = percent) +
  scale_ins() +
  labs(x = "Enrollment window (months after DOB)", y = "% of birth cohort",
       colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Each category against its own 12-month total. A curve rising later is a group
whose infants take longer to appear.

## Just those who had enrollment within 12 Months

```{r}
#| label: fig-enroll-normalized

enroll_pooled |>
  ggplot(aes(x = months, y = pct_of_group_final, colour = ins)) +
  geom_line() +
  geom_point(size = 1) +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(labels = percent) +
  scale_ins() +
  labs(x = "Enrollment window (months after DOB)",
       y = "% of category 12-month total", colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

`n_spine` is the whole birth cohort, so it is the same for every insurance row
in a year. The all-years row covers the complete cohorts only.

```{r}
#| label: table-enroll-window

enroll_wide <- enroll_sweep |>
  filter(months %in% c(1, 3, 6, 12)) |>
  select(birth_year, ins, n_spine, months, n_enrolled) |>
  pivot_wider(names_from = months, values_from = n_enrolled,
              names_glue = "{months}mo")

enroll_all_years <- enroll_sweep |>
  filter(birth_year <= last_complete_birth_year, months %in% c(1, 3, 6, 12)) |>
  group_by(ins, months) |>
  summarise(n_enrolled = sum(n_enrolled), .groups = "drop") |>
  pivot_wider(names_from = months, values_from = n_enrolled,
              names_glue = "{months}mo") |>
  mutate(birth_year = complete_years_label, n_spine = spine_complete)

bind_rows(enroll_wide |> mutate(birth_year = as.character(birth_year)),
          enroll_all_years) |>
  select(birth_year, ins, n_spine, `1mo`, `3mo`, `6mo`, `12mo`) |>
  mutate(across(where(is.numeric), ~ comma(.x, accuracy = 1))) |>
  ft("Infants clearing the enrollment requirement at 1, 3, 6 and 12 months")
```

## Time to first enrollment (those enrolled within a year)

Days from DOB to the start of the closest coverage segment.
Segments start on the first of the month, so birth-month coverage lands in bin 0.

```{r}
#| label: pull-enroll-days

enroll_days <- dbGetQuery(uthealth, glue("
  SELECT ins,
         LEAST(days_to_enroll, 400) AS days,
         COUNT(*) AS n
  FROM colour.infant_records
  WHERE days_to_enroll IS NOT NULL
    AND birth_year BETWEEN {year_min} AND {last_complete_birth_year}
  GROUP BY 1, 2
")) |> as_tibble()
```

```{r}
#| label: fig-enroll-days

enroll_days |>
  filter(days <= 360) |>
  mutate(bin = (days %/% 30) * 30) |>
  group_by(ins, bin) |>
  summarise(n = sum(n), .groups = "drop") |>
  ggplot(aes(x = bin, y = n)) +
  geom_col(width = 27) +
  facet_wrap(~ ins, scales = "free_y") +
  scale_x_continuous(breaks = seq(0, 360, 60)) +
  scale_y_continuous(labels = comma) +
  labs(x = "Days from DOB to enrollment", y = "Infants") +
  theme_minimal()
```

## Newborn-code claims

Among infants who cleared the enrollment requirement, the share with a
newborn-code claim within each window.

```{r}
#| label: pull-claim-sweep

claim_sweep_raw <- dbGetQuery(uthealth, glue("
  SELECT m.months,
         w.days AS claim_window,
         r.birth_year,
         r.ins,
         SUM(CASE WHEN r.days_to_enroll <= m.months * 30 THEN 1 ELSE 0 END) AS n_enrolled,
         SUM(CASE WHEN r.days_to_enroll <= m.months * 30
                   AND r.days_to_first_nb_claim <= w.days THEN 1 ELSE 0 END) AS n_with_claim
  FROM colour.infant_records r
  CROSS JOIN generate_series(1, 12) AS m(months)
  CROSS JOIN (VALUES (7), (30), (60), (90)) AS w(days)
  WHERE r.birth_year BETWEEN {year_min} AND {year_max}
  GROUP BY 1, 2, 3, 4
")) |> as_tibble()

claim_sweep <- claim_sweep_raw |>
  filter(ins != "Not Enrolled") |>
  add_ins_total(keys = c("months", "claim_window", "birth_year"),
                value_cols = c("n_enrolled", "n_with_claim")) |>
  mutate(pct_with_claim = n_with_claim / n_enrolled)

claim_pooled <- claim_sweep |>
  filter(birth_year <= last_complete_birth_year) |>
  group_by(months, claim_window, ins) |>
  summarise(across(c(n_enrolled, n_with_claim), sum), .groups = "drop") |>
  mutate(pct_with_claim = n_with_claim / n_enrolled)
```

Claim capture by window, at a `r enroll_window_months`-month enrollment window.

```{r}
#| label: fig-claim-window

claim_pooled |>
  filter(months == enroll_window_months) |>
  ggplot(aes(x = claim_window, y = pct_with_claim, colour = ins)) +
  geom_line() +
  geom_point(size = 1) +
  scale_x_continuous(breaks = c(7, 30, 60, 90)) +
  scale_y_continuous(labels = percent) +
  scale_ins() +
  labs(x = "Claim window (days after DOB)", y = "% of enrolled infants",
       colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Cumulative capture by day, with the four windows marked.

```{r}
#| label: pull-claim-cdf

claim_cdf <- dbGetQuery(uthealth, glue("
  SELECT ins,
         LEAST(days_to_first_nb_claim, 400) AS days,
         COUNT(*) AS n
  FROM colour.infant_records
  WHERE days_to_first_nb_claim IS NOT NULL
    AND days_to_enroll <= {enroll_window_days}
    AND birth_year BETWEEN {year_min} AND {last_complete_birth_year}
  GROUP BY 1, 2
")) |> as_tibble()

claim_denom <- claim_pooled |>
  filter(months == enroll_window_months, claim_window == 7) |>
  select(ins, n_enrolled)
```

```{r}
#| label: fig-claim-cdf

claim_cdf |>
  add_ins_total(keys = "days", value_cols = "n") |>
  left_join(claim_denom, by = "ins") |>
  arrange(ins, days) |>
  group_by(ins) |>
  mutate(cum_pct = cumsum(n) / first(n_enrolled)) |>
  ungroup() |>
  filter(days <= 120) |>
  ggplot(aes(x = days, y = cum_pct, colour = ins)) +
  geom_step() +
  geom_vline(xintercept = c(7, 30, 60, 90), linetype = "dotted", linewidth = 0.3) +
  scale_y_continuous(labels = percent) +
  scale_ins() +
  labs(x = "Days from DOB to first newborn-code claim",
       y = "Cumulative % of enrolled infants", colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Infant counts

`n_enrolled_in_window` is infants whose coverage began within
`r enroll_window_months` months of the DOB, not everyone who ever enrolled. The
ever-enrolled figure is in the waterfall table above.

```{r}
#| label: infant-headline

infant_counts <- claim_sweep_raw |>
  filter(months == enroll_window_months, claim_window == nb_claim_window_days) |>
  select(year = birth_year, ins, n_enrolled, n_with_claim) |>
  fold_not_enrolled(keys = "year", value_cols = c("n_enrolled", "n_with_claim")) |>
  add_ins_total(keys = "year", value_cols = c("n_enrolled", "n_with_claim")) |>
  rename(n_enrolled_in_window = n_enrolled) |>
  mutate(pct_with_claim = n_with_claim / n_enrolled_in_window)

infant_counts |>
  mutate(across(c(n_enrolled_in_window, n_with_claim), ~ comma(.x, accuracy = 1)),
         pct_with_claim = percent(pct_with_claim, accuracy = 0.1)) |>
  ft(glue("Infants at {enroll_window_months}-month enrollment / ",
          "{nb_claim_window_days}-day claim window"))
```

# Deliveries

## Attrition


```{r}
#| label: pull-delivery-attrition

delivery_attrition <- dbGetQuery(uthealth, "
  SELECT 1 AS step_num,
         '1. Delivery code hits (any age, incl. infant claims)' AS step,
         COUNT(DISTINCT pers_id) AS n_persons
  FROM colour.delivery_code_hits
  UNION ALL
  SELECT 2, '2. Joined to claim dates in medical_adj',
         COUNT(DISTINCT pers_id) FROM colour.delivery_claims_stg
  UNION ALL
  SELECT 3, '3. Age 12-50 at service, or no DOB on file',
         COUNT(DISTINCT pers_id) FROM colour.delivery_claims_qual
  UNION ALL
  SELECT 4, '4. Collapsed into an episode',
         COUNT(DISTINCT pers_id) FROM colour.delivery_episodes
  UNION ALL
  SELECT 5, '5. Episode dated within the analysis years',
         COUNT(DISTINCT pers_id) FROM colour.delivery_episodes_insurance_stg
  ORDER BY 1
") |> as_tibble()

delivery_attrition |>
  mutate(dropped = lag(n_persons) - n_persons,
         pct_kept = n_persons / first(n_persons)) |>
  mutate(across(c(n_persons, dropped), ~ comma(.x, accuracy = 1)),
         pct_kept = percent(pct_kept, accuracy = 0.1)) |>
  select(step, n_persons, dropped, pct_kept) |>
  ft("Distinct persons at each delivery pipeline step")
```

## Episode dating

Episodes anchored on a CPT date rather than falling back to a diagnosis date.
The date source drives the year assignment, so a large diagnosis share weakens
the year-by-year counts.

```{r}
#| label: table-dt-src

dbGetQuery(uthealth, "
  SELECT delivery_year AS year, delivery_dt_src, COUNT(*) AS n_episodes
  FROM colour.delivery_episodes_insurance_stg
  GROUP BY 1, 2 ORDER BY 1, 2
") |>
  as_tibble() |>
  pivot_wider(names_from = delivery_dt_src, values_from = n_episodes,
              values_fill = 0) |>
  mutate(pct_cpt = cpt / (cpt + diagnosis)) |>
  mutate(across(c(cpt, diagnosis), ~ comma(.x, accuracy = 1)),
         pct_cpt = percent(pct_cpt, accuracy = 0.1)) |>
  ft("Episode delivery date source by year")
```

## Delivery counts

One instance per person per year, taking the maximum `baby_count`. `n_mothers`
counts people; `n_babies` applies the baby count, so twins contribute 2.

```{r}
#| label: pull-delivery-counts

delivery_raw <- dbGetQuery(uthealth, glue("
  WITH person_year AS (
    SELECT pers_id,
           delivery_year,
           MAX(ins)        AS ins,
           MAX(baby_count) AS baby_count,
           COUNT(*)        AS n_episodes
    FROM colour.delivery_episodes_insurance_stg
    GROUP BY pers_id, delivery_year
  )
  SELECT delivery_year AS year,
         ins,
         COUNT(*)        AS n_mothers,
         SUM(baby_count) AS n_babies,
         SUM(n_episodes) AS n_episodes
  FROM person_year
  WHERE delivery_year BETWEEN {year_min} AND {year_max}
  GROUP BY 1, 2 ORDER BY 1, 2
")) |> as_tibble()

delivery_counts <- delivery_raw |>
  fold_not_enrolled(keys = "year", value_cols = c("n_mothers", "n_babies", "n_episodes")) |>
  add_ins_total(keys = "year", value_cols = c("n_mothers", "n_babies", "n_episodes"))

delivery_counts |>
  mutate(across(c(n_mothers, n_babies, n_episodes), ~ comma(.x, accuracy = 1))) |>
  ft("Deliveries by year and insurance, after the person-year collapse")
```

```{r}
#| label: pull-delivery-counts-raw

delivery_ep_raw <- dbGetQuery(uthealth, glue("
  SELECT delivery_year   AS year,
         ins,
         COUNT(*)        AS n_episodes_ep,
         SUM(baby_count) AS n_babies_ep
  FROM colour.delivery_episodes_insurance_stg
  WHERE delivery_year BETWEEN {year_min} AND {year_max}
  GROUP BY 1, 2 ORDER BY 1, 2
")) |> as_tibble()

delivery_ep_counts <- delivery_ep_raw |>
  fold_not_enrolled(keys = "year", value_cols = c("n_episodes_ep", "n_babies_ep")) |>
  add_ins_total(keys = "year", value_cols = c("n_episodes_ep", "n_babies_ep"))
```


## Length of stay versus CPT span

Three measures of how long the delivery event lasted:

- LOS: `discharge_dt - admit_dt`, only where an institutional claim is present.
- CPT span: `cpt_dos_max - cpt_dos_min`, how far the procedure lines spread.
- Episode length: `episode_end - episode_start`, the full derived span.


```{r}
#| label: pull-los

los_marginals <- dbGetQuery(uthealth, "
  SELECT 'LOS (discharge - admit)' AS measure,
         LEAST(discharge_dt - admit_dt, 21) AS days,
         COUNT(*) AS n
  FROM colour.delivery_episodes_insurance_stg
  WHERE admit_dt IS NOT NULL AND discharge_dt IS NOT NULL
  GROUP BY 1, 2
  UNION ALL
  SELECT 'CPT span (max - min)',
         LEAST(cpt_dos_max - cpt_dos_min, 21),
         COUNT(*)
  FROM colour.delivery_episodes_insurance_stg
  WHERE cpt_dos_min IS NOT NULL AND cpt_dos_max IS NOT NULL
  GROUP BY 1, 2
  UNION ALL
  SELECT 'Episode length',
         LEAST(episode_len_days, 21),
         COUNT(*)
  FROM colour.delivery_episodes_insurance_stg
  GROUP BY 1, 2
") |> as_tibble()

cpt_position <- dbGetQuery(uthealth, "
  SELECT CASE
           WHEN cpt_dos_min < admit_dt     THEN 'CPT before admission'
           WHEN cpt_dos_min > discharge_dt THEN 'CPT after discharge'
           ELSE 'CPT within the stay'
         END AS position,
         COUNT(*) AS n_episodes
  FROM colour.delivery_episodes_insurance_stg
  WHERE admit_dt IS NOT NULL AND discharge_dt IS NOT NULL
    AND cpt_dos_min IS NOT NULL
  GROUP BY 1 ORDER BY 2 DESC
") |> as_tibble()
```

Distribution of the three measures, capped at 21 days.

```{r}
#| label: fig-los-marginals
#| fig-height: 5.5

los_marginals |>
  group_by(measure) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  filter(days >= 0) |>
  ggplot(aes(x = days, y = pct)) +
  geom_col(width = 0.85) +
  facet_wrap(~ measure, ncol = 1, scales = "free_y") +
  scale_x_continuous(breaks = seq(0, 21, 3)) +
  scale_y_continuous(labels = percent) +
  labs(x = "Days (capped at 21)", y = "% of episodes with the measure") +
  theme_minimal()
```

```{r}
#| label: table-cpt-position

cpt_position |>
  mutate(pct = percent(n_episodes / sum(n_episodes), accuracy = 0.1),
         n_episodes = comma(n_episodes, accuracy = 1)) |>
  ft("Where the first CPT date falls relative to the recorded admission")
```



# Delivery data quality

## Pregnancies per person


```{r}
#| label: pull-preg-per-person

preg_per_person <- dbGetQuery(uthealth, "
  SELECT n_ep, COUNT(*) AS n_persons FROM (
    SELECT pers_id, COUNT(*) AS n_ep
      FROM colour.delivery_episodes_insurance_stg
     GROUP BY 1) t
  GROUP BY 1 ORDER BY 1
") |> as_tibble()

```

```{r}
#| label: fig-preg-per-person

preg_per_person |>
  mutate(n_ep_lab = if_else(n_ep >= 6, "6+", as.character(n_ep))) |>
  group_by(n_ep_lab) |>
  summarise(n_persons = sum(n_persons), .groups = "drop") |>
  mutate(n_ep_lab = factor(n_ep_lab, levels = c(as.character(1:5), "6+"))) |>
  ggplot(aes(x = n_ep_lab, y = n_persons)) +
  geom_col(width = 0.7) +
  scale_y_continuous(labels = comma) +
  labs(x = "Delivery episodes per person", y = "Persons") +
  theme_minimal()
```

```{r}
#| label: table-preg-per-person

preg_per_person |>
  mutate(pct = percent(n_persons / sum(n_persons), accuracy = 0.1),
         n_persons = comma(n_persons, accuracy = 1)) |>
  head(15) |>
  ft("Delivery episodes per person (first 10 rows)")
```



## Gaps between pregnancies

Interval between consecutive `delivery_dt` values for people with more than one
episode. 


```{r}
#| label: pull-preg-gaps

preg_gaps <- dbGetQuery(uthealth, "
  WITH ordered AS (
    SELECT pers_id,
           delivery_dt,
           LAG(delivery_dt) OVER (PARTITION BY pers_id ORDER BY delivery_dt) AS prev_dt
    FROM colour.delivery_episodes_insurance_stg
  )
  SELECT LEAST(delivery_dt - prev_dt, 1825) AS gap_days,
         COUNT(*) AS n
  FROM ordered
  WHERE prev_dt IS NOT NULL
  GROUP BY 1 ORDER BY 1
") |> as_tibble()
```

Full range, capped at five years.

```{r}
#| label: fig-preg-gaps

preg_gaps |>
  ggplot(aes(x = gap_days, y = n)) +
  geom_col(width = 1) +
  geom_vline(xintercept = min_preg_gap_days, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0, 1800, 180)) +
  scale_y_continuous(labels = comma) +
  labs(x = "Days between consecutive delivery dates", y = "Intervals") +
  theme_minimal()
```

First year only, where the implausible intervals sit.

```{r}
#| label: fig-preg-gaps-short

preg_gaps |>
  filter(gap_days <= 365) |>
  ggplot(aes(x = gap_days, y = n)) +
  geom_col(width = 1) +
  geom_vline(xintercept = min_preg_gap_days, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0, 365, 30)) +
  scale_y_continuous(labels = comma) +
  labs(x = "Days between consecutive delivery dates", y = "Intervals") +
  theme_minimal()
```

```{r}
#| label: table-preg-gaps

preg_gaps |>
  mutate(band = case_when(
    gap_days <  30 ~ "Under 30 days",
    gap_days <  90 ~ "30-89 days",
    gap_days < 180 ~ "90-179 days",
    gap_days < min_preg_gap_days ~ glue("180-{min_preg_gap_days - 1} days"),
    TRUE ~ glue("{min_preg_gap_days}+ days")
  )) |>
  mutate(band = factor(band, levels = c("Under 30 days", "30-89 days", "90-179 days",
                                        glue("180-{min_preg_gap_days - 1} days"),
                                        glue("{min_preg_gap_days}+ days")))) |>
  group_by(band) |>
  summarise(n_intervals = sum(n), .groups = "drop") |>
  mutate(pct = percent(n_intervals / sum(n_intervals), accuracy = 0.1),
         n_intervals = comma(n_intervals, accuracy = 1)) |>
  ft("Intervals between consecutive delivery dates, banded")
```

# If you only include CPT deliveries, does the weird gaps go away? 

```{r}
#| label: pull-preg-gaps-cpt

preg_gaps_cpt <- dbGetQuery(uthealth, "
  WITH ordered AS (
    SELECT pers_id,
           delivery_dt,
           LAG(delivery_dt) OVER (PARTITION BY pers_id ORDER BY delivery_dt) AS prev_dt
    FROM colour.delivery_episodes_insurance_stg
    WHERE delivery_dt_src = 'cpt'
  )
  SELECT LEAST(delivery_dt - prev_dt, 1825) AS gap_days,
         COUNT(*) AS n
  FROM ordered
  WHERE prev_dt IS NOT NULL
  GROUP BY 1 ORDER BY 1
") |> as_tibble()

```

```{r}
#| label: fig-preg-gaps-cpt

bind_rows(
  preg_gaps     |> mutate(series = "All episodes"),
  preg_gaps_cpt |> mutate(series = "CPT-dated only")
) |>
  group_by(series) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  filter(gap_days <= 365) |>
  mutate(bin = (gap_days %/% 7) * 7) |>
  group_by(series, bin) |>
  summarise(pct = sum(pct), .groups = "drop") |>
  ggplot(aes(x = bin, y = pct, colour = series)) +
  geom_line() +
  geom_vline(xintercept = min_preg_gap_days, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0, 365, 30)) +
  scale_y_continuous(labels = percent) +
  labs(x = "Days between consecutive delivery dates (7-day bins)",
       y = "% of all intervals in the series", colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


# Vital statistics 


```{r}
#| label: pull-vital

vital_long <- dbGetQuery(uthealth, glue("
  SELECT * FROM colour.ref_vital_stats_tx
  WHERE year BETWEEN {year_min} AND {year_max}
  ORDER BY year
")) |>
  as_tibble() |>
  transmute(
    year,
    Medicaid   = medicaid,
    Commercial = private_insurance,
    Other      = self_pay + other + unknown_or_not_stated,
    Total      = total
  ) |>
  pivot_longer(c(Medicaid, Commercial, Other, Total),
               names_to = "ins", values_to = "vital_n")
```

```{r}
#| label: vital-comparison

vital_comparison <- bind_rows(
  delivery_counts |>
    transmute(pipeline = "Deliveries (babies)", year, ins = as.character(ins), n = n_babies),
  infant_counts |>
    transmute(pipeline = glue("Infants ({nb_claim_window_days}-day claim)"),
              year, ins = as.character(ins), n = n_with_claim)
) |>
  left_join(vital_long, by = c("year", "ins")) |>
  mutate(diff = n - vital_n,
         pct_diff = diff / vital_n,
         ins = factor(ins, levels = ins_levels)) |>
  arrange(pipeline, year, ins)

vital_comparison |>
  filter(ins == "Total") |>
  select(pipeline, year, n, vital_n, diff, pct_diff) |>
  mutate(across(c(n, vital_n, diff), ~ comma(.x, accuracy = 1)),
         pct_diff = percent(pct_diff, accuracy = 0.1)) |>
  ft("Claims counts against vital statistics births, all payers")
```

Percent difference from vital statistics, by category.

```{r}
#| label: fig-vital

vital_comparison |>
  filter(ins != "Total", !is.na(pct_diff)) |>
  ggplot(aes(x = year, y = pct_diff, colour = ins)) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_line() +
  geom_point(size = 1) +
  facet_wrap(~ pipeline) +
  scale_y_continuous(labels = percent) +
  scale_ins() +
  labs(x = NULL, y = "% difference from vital stats", colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Excel export


```{r}
#| label: export

# -- EDIT THIS ----------------------------------------------------
out_path <- here::here("infant_delivery_counts.xlsx")
# -----------------------------------------------------------------

sheets <- list(
  "Infant Waterfall" = waterfall,
  "Infant Counts"    = infant_counts,
  "Infant Windows"   = claim_sweep |>
                         select(months, claim_window, birth_year, ins,
                                n_enrolled, n_with_claim, pct_with_claim),
  "Enroll Windows"   = enroll_sweep |>
                         select(months, birth_year, ins, n_spine, n_enrolled,
                                pct_of_spine, pct_of_group_final),
  "Delivery Counts"  = delivery_counts,
  "vs Vital Stats"   = vital_comparison
)

wb <- createWorkbook()
pct_style <- createStyle(numFmt = "0.0%")

set_widths_for_headers <- function(wb, sheet, df) {
  if (nrow(df) == 0) return(invisible(NULL))
  col_widths <- vapply(seq_along(df), function(i) {
    header_len <- nchar(names(df)[i])
    data_len   <- suppressWarnings(max(nchar(format(df[[i]])), na.rm = TRUE))
    if (!is.finite(data_len)) data_len <- 0
    max(header_len, data_len) + 2
  }, numeric(1))
  setColWidths(wb, sheet, cols = seq_along(df), widths = col_widths)
}

for (sheet_name in names(sheets)) {
  df <- sheets[[sheet_name]] |> mutate(across(where(is.factor), as.character))

  addWorksheet(wb, sheet_name)
  writeDataTable(wb, sheet_name, df, tableStyle = "TableStyleMedium9")

  pct_cols <- which(grepl("^pct_", names(df)))
  if (length(pct_cols) > 0 && nrow(df) > 0) {
    addStyle(wb, sheet_name, pct_style,
             rows = 2:(nrow(df) + 1), cols = pct_cols, gridExpand = TRUE)
  }

  set_widths_for_headers(wb, sheet_name, df)
}

notes_text <- c(
  "Infant and delivery counts - window sensitivity",
  "",
  glue("Headline windows: enrollment {enroll_window_months} months, ",
       "newborn-code claim {nb_claim_window_days} days"),
  "",
  "Infants (applied in sequence, neither materialised in SQL):",
  "- Spine = every person in colour.dob with a single dob on/after 2019-01-01",
  "- Step 1: enrollment begins within the enrollment window of the DOB",
  "- Step 2: a newborn-code claim appears within the claim window of the DOB,",
  "  among infants who cleared step 1",
  "- n_enrolled_in_window is timing-based, not ever-enrolled; the ever-enrolled",
  "  count is in the Infant Waterfall tab",
  "- Enrollment = the coverage segment closest to the DOB; a segment spanning",
  "  the DOB scores 0 days. Ties go to Commercial",
  "- A window month is 30 days, so 12 months = 360 (SUMMARY B in the SQL uses 365)",
  "- Insurance is the plan on the picked segment, not agg_yr_plan",
  glue("- Birth years after {last_complete_birth_year} lack full follow-up and are ",
       "censored, not low. Figures pool complete cohorts only"),
  "",
  "Deliveries:",
  "- Delivery Codes tab of the Orr codelists (CPT and DX; no ICD-10-PCS codes)",
  "- Delivery and newborn code lists overlap, so infants carrying a delivery-tab",
  "  code are picked up at step 1 of the attrition and removed by the age filter",
  "- Persons kept where any matched claim falls at age 12-50, or where no DOB",
  "  is on file",
  "- Overlapping claim date spans collapsed into pseudo-episodes; the merge uses",
  "  the full claim span, not the CPT dates, so dx-only claims can join",
  "- Episode delivery date = first CPT line date, falling back to episode start",
  "- Person-year collapse (max baby_count) applied in the analysis, not the SQL",
  "- Insurance is prim_med_plan from agg_yr_plan for the delivery year",
  "",
  "Notes:",
  "- Not Enrolled is folded into Other to match the 3-way vital stats split",
  "- Vital stats record the mother's payer, not the baby's",
  "- Data quality checks are in the Word document"
)

addWorksheet(wb, "Notes")
writeData(wb, "Notes", notes_text, colNames = FALSE)
setColWidths(wb, "Notes", cols = 1, widths = 95)

saveWorkbook(wb, out_path, overwrite = TRUE)
cat("Wrote", out_path, "\n")
```

