---
title: "Build Reference Table: Orr 2024 Codelists"
format: html
---

## Overview

Reads every sheet in `Orr2024_Codelists.xlsx` (Delivery, Delivery Twin Excl.,
Newborn, Newborn Twin Excl., Stillbirth Excl.), stacks them into a single long
table, cleans column names with `janitor`, and writes the result to
`colour.ref_orr2024_codelists` on Greenplum.

Each sheet already carries a `Tab` column identifying its source list, so the
stacked table is self-labeling - no extra join key needed.

The `Delivery` and `Delivery Twin Excl.` sheets now also carry a `Baby Count`
column (number of infants implied by that ICD-10-CM code, sourced from
BabyCounts.xlsx). It's only populated for ICD-10-CM diagnosis rows on those
two sheets - CPT/ICD-10-PCS procedure codes and the other three sheets don't
have a baby count and stack in as `NA`. `map_dfr()` handles the ragged
columns fine, but since it's read in as text like everything else, it gets
coerced to integer after cleaning so it lands as a numeric column on
Greenplum instead of text.

```{r}
#| label: setup
#| message: false

pacman::p_load(RPostgres, DBI, keyring, janitor, tidyverse, readxl, stringi)

current_dir <- here::here()
```

```{r}
#| label: params

# -- EDIT THESE ----------------------------------------------
# Codelist file lives one directory up from the project root, in /data
xlsx_path    <- here::here("..", "data", "Orr2024_Codelists_with_baby_count.xlsx")
target_schema <- "colour"
target_table  <- "ref_orr2024_codelists"
# ------------------------------------------------------------
```

```{r}
#| label: read-and-stack

sheet_names <- excel_sheets(xlsx_path)

codelists <- map_dfr(
  sheet_names,
  ~ read_excel(xlsx_path, sheet = .x, col_types = "text") |>
    mutate(source_sheet = .x)
) |>
  clean_names()

glimpse(codelists)
count(codelists, tab)
count(codelists, tab, has_baby_count = !is.na(baby_count))
```

```{r}
#| label: clean-funky-characters

clean_text <- function(x) {
  x |>
    stri_trans_general("Any-Latin; Latin-ASCII") |>  # smart quotes, dashes, accents -> ASCII
    str_replace_all("[\u00A0\u2007\u202F]", " ") |>   # non-breaking / narrow spaces -> regular space
    str_replace_all("\\s+", " ") |>                   # collapse repeated whitespace
    str_trim()
}

codelists <- codelists |>
  mutate(across(where(is.character), clean_text))

glimpse(codelists)
```

```{r}
#| label: type-baby-count

# baby_count comes in as text (col_types = "text" on the read) - coerce to
# integer so it writes to Greenplum as a numeric column, not text. Compare
# NA counts before/after so a stray non-numeric value (e.g. a typo'd cell)
# fails loudly instead of silently becoming NA.
na_before <- sum(is.na(codelists$baby_count))

codelists <- codelists |>
  mutate(baby_count = as.integer(baby_count))

na_after <- sum(is.na(codelists$baby_count))

stopifnot(
  "as.integer() introduced new NAs in baby_count - check for non-numeric values" =
    na_after == na_before
)

count(codelists, tab, baby_count)
```

```{r}
#| label: connect

# Same Greenplum server/credentials as the other pipelines (apcd-paper2,
# POEM TX paper1 v24) - different database (uthealth).
greenplum_user <- "jwozny"

uthealth <- dbConnect(
  RPostgres::Postgres(),
  dbname   = "apcd_dev",
  user     = greenplum_user,
  password = keyring::key_get("apcd_dev", greenplum_user),
  host     = "spgpcdplv001",
  bigint   = "integer"
)
```



```{r}
#| label: write-table

dbWriteTable(
  uthealth,
  Id(schema = target_schema, table = target_table),
  codelists,
  overwrite = TRUE
)

cat(
  "Wrote", nrow(codelists), "rows to",
  paste0(target_schema, ".", target_table), "\n"
)
```

```{r}
#| label: verify

check <- dbGetQuery(
  uthealth,
  glue::glue_sql(
    "SELECT tab,
            count(*) AS n,
            count(baby_count) AS n_with_baby_count
     FROM {`target_schema`}.{`target_table`}
     GROUP BY tab
     ORDER BY tab",
    .con = uthealth
  )
)
check
```

```{r}
#| label: disconnect

dbDisconnect(uthealth)
```
