---
title: "Delivery & Newborn Claims Counts vs. TX Vital Statistics"
format: html
---

## Overview

Compares two independent claims-based counts against `colour.ref_vital_stats_tx`
(CDC WONDER TX birth counts by source of payment for delivery), each of which
should approximate vital-stats **births** in its own way:

- **Deliveries** (`colour.delivery_records_insurance_stg`, built from the
  `Delivery` tab of the Orr 2024 codelist) - a person-year-grain
  staging table. Two counts are pulled from it, since a delivery episode and
  a birth aren't the same thing once twins are in the mix:
  - **`n_mothers`** - `COUNT(DISTINCT pers_id)`, i.e. distinct people with a
    qualifying delivery in that year
  - **`n_babies`** - the same deliveries, summed by `baby_count`, so a set of
    twins contributes 2. This is the count that should line up with vital
    stats births, while `n_mothers` will run slightly below it.
- **Newborns** (`colour.newborn_records_insurance_stg`, built from the
  `newborn` tab of the codelist) - a person-year-grain staging
  table. Only one count applies here:
  - **`n_newborns`** - `COUNT(DISTINCT pers_id)`, i.e. distinct newborns
    identified in claims that year

The enrollment age filters (`age > 12` for deliveries, `age < 1` for newborns)
have been **removed** from both staging tables - the DOB criteria at stage 2 are
now the only age control. Two consequences worth holding onto:

1. **Stage 0 is looser than the old pipeline**, not equal to it. It is now every
   person with a matched code in that year at any age, so it will run above the
   counts these tables used to report. Stages 2-3 are the comparable numbers.
2. **`Not Enrolled` is now a populated bucket.** The old age filters were
   NULL-unsafe, so anyone without an `agg_yr_plan` match for that year was
   silently dropped; without them those people survive with a NULL
   `prim_med_plan`. They are folded into `Other` to match the 3-way vital stats
   split, which means `Other` now carries genuinely unenrolled people on top of
   Medicare/Federal. The *Enrollment QC* section sizes that before folding.

Both pipelines assign insurance the same way - `prim_med_plan` from
`research_di.agg_yr_plan` for that year (an enrollment-based check) - so the
`ins` categories line up 1:1 across deliveries, newborns, and vital stats:

| Claims (`ins`) | `prim_med_plan` values                                                              | Vital stats                                  |
|-----------------|--------------------------------------------------------------------------------------|-----------------------------------------------|
| Medicaid         | Medicaid                                                                              | `medicaid`                                     |
| Commercial       | Commercial, Com-Ers, Com-ErsTrs, Com-Trs                                              | `private_insurance`                            |
| Other            | Medicare Advantage(+ Imputed), Medicare FFS, Medicare Imputed, Federal, Not Enrolled  | `self_pay` + `other` + `unknown_or_not_stated` |
| Total            | (all of the above)                                                                    | `total`                                        |

Both tables below carry a per-year **Total** row (summed across insurance
categories) alongside the by-category breakdown, and both report a raw
difference (`claims - vital`) and a % difference (`diff / vital`) so
over/under-capture is easy to spot at a glance.

### DOB linkage and filter stages

`colour.dob` is LEFT JOINed onto both claim-level staging tables, so every
matched claim carries the person's derived DOB (`dob_derv`) and the multiple-DOB
flag. Nothing is dropped by that join - instead each pipeline materializes a
**cumulative filter waterfall** (`colour.delivery_counts_by_stage`,
`colour.newborn_counts_by_stage`):

| Stage | Deliveries (mothers)              | Newborns (infants)                |
|-------|-----------------------------------|-----------------------------------|
| 0     | All matched claims (pre-DOB)      | All matched claims (pre-DOB)      |
| 1     | + DOB present                     | + DOB present                     |
| 2     | + claim with age 12-50 at DOS     | + claim within 7 days of DOB      |
| 3     | + single DOB only                 | + single DOB only                 |

**The headline tables and figures below report the fully filtered population
(stage 3).** Stage 0 - the pre-DOB numbers - and everything in between are
in the *Delivery Stages*, *Newborn Stages*, *Claim Counts*, and *Stages vs
Vital* tabs, so the attrition and the original counts stay one click away
rather than being the default view. `headline_stage` in the setup below is the
single place to change that.

From stage 2 on, claim counts (and, for deliveries, `n_babies`) are restricted
to qualifying claims only - `n_babies_asis` in the stage tables carries the
unrestricted `baby_count` through every stage for comparison.
`colour.*_dob_qc` splits the stage-0 population into No DOB / Single DOB /
Multiple DOB so each exclusion can also be sized independently of the order
they're applied in.

```{r}
#| label: setup
#| message: false

pacman::p_load(RPostgres, DBI, keyring, glue, tidyverse, scales, gt, openxlsx)
```

```{r}
#| label: connect

greenplum_user <- "jwozny"

uthealth <- dbConnect(
  RPostgres::Postgres(),
  dbname   = "apcd_dev",
  user     = greenplum_user,
  password = keyring::key_get("apcd_dev", greenplum_user),
  host     = "spgpcdplv001",
  bigint   = "integer"
)
```

```{r}
#| label: headline-stage

# -- EDIT THIS ----------------------------------------------------
# Which filter stage drives the headline tables (Overall, Deliveries,
# Newborns) and the Excel tabs built from them:
#   0 = all matched claims (pre-DOB behaviour)
#   1 = + DOB present
#   2 = + qualifying claim (7-day window / age 12-50)
#   3 = + single DOB only            <- fully filtered
# The stage tabs always carry all four stages regardless.
headline_stage <- 3
# -----------------------------------------------------------------
```

```{r}
#| label: pull-vital-stats

vital_stats <- dbGetQuery(uthealth, "
  SELECT * FROM colour.ref_vital_stats_tx
  WHERE year >= 2019
  ORDER BY year
") |>
  as_tibble()

vital_stats
```

```{r}
#| label: reshape-vital-stats

# Collapse to the same 4 buckets (3 insurance categories + Total) that both
# claims-side tables use, so joins below are a straight (year, ins) match.
vital_long <- vital_stats |>
  transmute(
    year,
    Medicaid   = medicaid,
    Commercial = private_insurance,
    Other      = self_pay + other + unknown_or_not_stated,
    Total      = total
  ) |>
  pivot_longer(
    cols = c(Medicaid, Commercial, Other, Total),
    names_to  = "ins",
    values_to = "vital_n"
  )

vital_long
```

```{r}
#| label: ins-order

# Shared factor levels so both output tables sort Medicaid / Commercial /
# Other / Total consistently within each year.
ins_levels <- c("Medicaid", "Commercial", "Other", "Total")
```

```{r}
#| label: helpers

# Both pipelines need the same two shape-fixing steps, and the stage tables
# need them once per stage, so they're factored out here.

# Fold the QC-only 'Not Enrolled' bucket into 'Other' so the categories line
# up 1:1 with the 3-way vital stats split, then re-aggregate.
fold_not_enrolled <- function(df, keys, value_cols) {
  df |>
    mutate(ins = if_else(ins == "Not Enrolled", "Other", ins)) |>
    group_by(across(all_of(c(keys, "ins")))) |>
    summarise(across(all_of(value_cols), sum), .groups = "drop")
}

# Append a per-key 'Total' row summed across insurance categories.
add_ins_total <- function(df, keys, value_cols) {
  totals <- df |>
    group_by(across(all_of(keys))) |>
    summarise(across(all_of(value_cols), sum), .groups = "drop") |>
    mutate(ins = "Total")

  bind_rows(df, totals) |>
    mutate(ins = factor(ins, levels = ins_levels)) |>
    # !!!syms rather than across()/pick() so this works on older dplyr too
    arrange(!!!syms(c(keys, "ins")))
}
```

## Pulling the stage tables

Everything downstream - headline tables included - is derived from these two
frames, so the filtered and unfiltered views can never drift apart: they are
literally different rows of the same pull.

```{r}
#| label: pull-stage-counts

delivery_value_cols <- c("n_mothers", "n_babies", "n_babies_asis", "n_claims", "n_code_hits")
newborn_value_cols  <- c("n_newborns", "n_claims", "n_code_hits")

# The _raw frames keep 'Not Enrolled' as its own category. Now that the
# enrollment age filters are gone that bucket is actually populated, so it is
# worth measuring before it gets folded into 'Other' (see Enrollment QC below).
delivery_stages_raw <- dbGetQuery(uthealth, "
  SELECT stage_num, stage, year, ins,
         n_mothers, n_babies, n_babies_asis, n_claims, n_code_hits
  FROM colour.delivery_counts_by_stage
  ORDER BY stage_num, year, ins
") |>
  as_tibble()

newborn_stages_raw <- dbGetQuery(uthealth, "
  SELECT stage_num, stage, year, ins,
         n_newborns, n_claims, n_code_hits
  FROM colour.newborn_counts_by_stage
  ORDER BY stage_num, year, ins
") |>
  as_tibble()

delivery_stages <- delivery_stages_raw |>
  fold_not_enrolled(keys = c("stage_num", "stage", "year"),
                    value_cols = delivery_value_cols) |>
  add_ins_total(keys = c("stage_num", "stage", "year"),
                value_cols = delivery_value_cols)

newborn_stages <- newborn_stages_raw |>
  fold_not_enrolled(keys = c("stage_num", "stage", "year"),
                    value_cols = newborn_value_cols) |>
  add_ins_total(keys = c("stage_num", "stage", "year"),
                value_cols = newborn_value_cols)

# Label for the selected headline stage, pulled from the data rather than
# hard-coded so table titles can't fall out of sync with headline_stage.
headline_label <- delivery_stages |>
  filter(stage_num == headline_stage) |>
  pull(stage) |>
  unique()

stopifnot(length(headline_label) == 1)
headline_label

delivery_stages
newborn_stages
```

## Deliveries (Mothers + Babies)

```{r}
#| label: delivery-headline

# Headline = the selected stage only. n_babies here is the stage-restricted
# birth count (age-eligible claims only from stage 2 on); n_babies_asis for
# the same population is in the Delivery Stages tab.
delivery_all <- delivery_stages |>
  filter(stage_num == headline_stage) |>
  select(year, ins, n_mothers, n_babies)

delivery_all
```

```{r}
#| label: delivery-compare

delivery_comparison <- delivery_all |>
  left_join(vital_long, by = c("year", "ins")) |>
  mutate(
    diff_mothers     = n_mothers - vital_n,
    pct_diff_mothers = diff_mothers / vital_n,
    diff_babies      = n_babies - vital_n,
    pct_diff_babies  = diff_babies / vital_n
  ) |>
  arrange(year, ins)

delivery_comparison
```

```{r}
#| label: table-deliveries

delivery_comparison |>
  mutate(
    pct_diff_mothers = percent(pct_diff_mothers, accuracy = 0.1),
    pct_diff_babies  = percent(pct_diff_babies, accuracy = 0.1)
  ) |>
  gt(groupname_col = "year") |>
  tab_header(
    title    = "Claims-Identified Deliveries vs. TX Vital Statistics Births",
    subtitle = glue("After DOB filtering - stage {headline_stage}: {headline_label}")
  ) |>
  cols_label(
    ins               = "Insurance Category",
    n_mothers         = "Claims Mothers",
    n_babies          = "Claims Births (baby_count-adj.)",
    vital_n           = "Vital Stats Births",
    diff_mothers      = "Diff (Mothers)",
    pct_diff_mothers  = "% Diff (Mothers)",
    diff_babies       = "Diff (Births)",
    pct_diff_babies   = "% Diff (Births)"
  ) |>
  fmt_number(columns = c(n_mothers, n_babies, vital_n, diff_mothers, diff_babies), decimals = 0)
```

```{r}
#| label: fig-deliveries
#| fig-cap: "Claims-identified births (baby_count-adjusted, after DOB filtering) vs. vital statistics births, by insurance category"

delivery_comparison |>
  filter(ins != "Total") |>
  pivot_longer(cols = c(n_babies, vital_n), names_to = "source", values_to = "n") |>
  mutate(source = recode(source, n_babies = "Claims (Births)", vital_n = "Vital Stats")) |>
  ggplot(aes(x = year, y = n, color = source)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ ins, scales = "free_y") +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Count", color = NULL) +
  theme_minimal()
```

## Newborns

```{r}
#| label: newborn-headline

newborn_all <- newborn_stages |>
  filter(stage_num == headline_stage) |>
  select(year, ins, n_newborns)

newborn_all
```

```{r}
#| label: newborn-compare

newborn_comparison <- newborn_all |>
  left_join(vital_long, by = c("year", "ins")) |>
  mutate(
    diff_newborns     = n_newborns - vital_n,
    pct_diff_newborns = diff_newborns / vital_n
  ) |>
  arrange(year, ins)

newborn_comparison
```

```{r}
#| label: table-newborns

newborn_comparison |>
  mutate(pct_diff_newborns = percent(pct_diff_newborns, accuracy = 0.1)) |>
  gt(groupname_col = "year") |>
  tab_header(
    title    = "Claims-Identified Newborns vs. TX Vital Statistics Births",
    subtitle = glue("After DOB filtering - stage {headline_stage}: {headline_label}")
  ) |>
  cols_label(
    ins               = "Insurance Category",
    n_newborns        = "Claims Newborns",
    vital_n           = "Vital Stats Births",
    diff_newborns     = "Diff (Newborns)",
    pct_diff_newborns = "% Diff (Newborns)"
  ) |>
  fmt_number(columns = c(n_newborns, vital_n, diff_newborns), decimals = 0)
```

```{r}
#| label: fig-newborns
#| fig-cap: "Claims-identified newborns (after DOB filtering) vs. vital statistics births, by insurance category"

newborn_comparison |>
  filter(ins != "Total") |>
  pivot_longer(cols = c(n_newborns, vital_n), names_to = "source", values_to = "n") |>
  mutate(source = recode(source, n_newborns = "Claims (Newborns)", vital_n = "Vital Stats")) |>
  ggplot(aes(x = year, y = n, color = source)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ ins, scales = "free_y") +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Count", color = NULL) +
  theme_minimal()
```

## Overall Summary

Combines just the per-year **Total** rows from both pipelines (all insurance
categories combined) into one high-level table, at the headline filter stage.

```{r}
#| label: overall-compare

overall_comparison <- delivery_comparison |>
  filter(ins == "Total") |>
  select(year, vital_n, n_mothers, diff_mothers, pct_diff_mothers,
         n_babies, diff_babies, pct_diff_babies) |>
  left_join(
    newborn_comparison |>
      filter(ins == "Total") |>
      select(year, n_newborns, diff_newborns, pct_diff_newborns),
    by = "year"
  )

overall_comparison
```

```{r}
#| label: table-overall

overall_comparison |>
  mutate(
    pct_diff_mothers  = percent(pct_diff_mothers, accuracy = 0.1),
    pct_diff_babies   = percent(pct_diff_babies, accuracy = 0.1),
    pct_diff_newborns = percent(pct_diff_newborns, accuracy = 0.1)
  ) |>
  gt() |>
  tab_header(
    title    = "Overall (All Insurance Categories Combined) vs. TX Vital Statistics Births",
    subtitle = glue("After DOB filtering - stage {headline_stage}: {headline_label}")
  ) |>
  cols_label(
    year              = "Year",
    vital_n           = "Vital Stats Births",
    n_mothers         = "Claims Mothers",
    diff_mothers      = "Diff (Mothers)",
    pct_diff_mothers  = "% Diff (Mothers)",
    n_babies          = "Claims Births (baby_count-adj.)",
    diff_babies       = "Diff (Births)",
    pct_diff_babies   = "% Diff (Births)",
    n_newborns        = "Claims Newborns",
    diff_newborns     = "Diff (Newborns)",
    pct_diff_newborns = "% Diff (Newborns)"
  ) |>
  fmt_number(
    columns = c(vital_n, n_mothers, diff_mothers, n_babies, diff_babies, n_newborns, diff_newborns),
    decimals = 0
  )
```

# DOB Filter Effects

Everything from here down carries all four stages, including the pre-DOB
stage 0 that the headline tables above used to report.

```{r}
#| label: stage-effect

# Within each (year, ins), express every stage relative to the one above it
# (`_drop`, negative = rows lost) and relative to stage 0 (`_pct_of_base`).
add_stage_effect <- function(df, value_cols) {
  df |>
    group_by(year, ins) |>
    arrange(stage_num, .by_group = TRUE) |>
    mutate(across(
      all_of(value_cols),
      list(
        drop        = ~ .x - lag(.x),
        pct_of_base = ~ .x / first(.x)
      ),
      .names = "{.col}_{.fn}"
    )) |>
    ungroup() |>
    arrange(year, ins, stage_num)
}

delivery_stage_effect <- delivery_stages |>
  add_stage_effect(c("n_mothers", "n_babies", "n_claims"))

newborn_stage_effect <- newborn_stages |>
  add_stage_effect(c("n_newborns", "n_claims"))

delivery_stage_effect
newborn_stage_effect
```

## Deliveries: what each exclusion costs

```{r}
#| label: table-delivery-stages

delivery_stage_effect |>
  filter(ins == "Total") |>
  select(year, stage, n_mothers, n_mothers_drop, n_mothers_pct_of_base,
         n_babies, n_babies_asis, n_claims, n_claims_drop, n_claims_pct_of_base) |>
  mutate(
    n_mothers_pct_of_base = percent(n_mothers_pct_of_base, accuracy = 0.1),
    n_claims_pct_of_base  = percent(n_claims_pct_of_base, accuracy = 0.1)
  ) |>
  gt(groupname_col = "year") |>
  tab_header(
    title    = "Deliveries: Effect of DOB-Based Exclusions",
    subtitle = "All insurance categories combined; cumulative stages. Stage 0 = pre-DOB counts"
  ) |>
  cols_label(
    stage                 = "Stage",
    n_mothers             = "Mothers",
    n_mothers_drop        = "Mothers Lost",
    n_mothers_pct_of_base = "% of Stage 0",
    n_babies              = "Births (stage-restricted)",
    n_babies_asis         = "Births (as is)",
    n_claims              = "Claims",
    n_claims_drop         = "Claims Lost",
    n_claims_pct_of_base  = "% of Stage 0"
  ) |>
  fmt_number(
    columns = c(n_mothers, n_mothers_drop, n_babies, n_babies_asis, n_claims, n_claims_drop),
    decimals = 0
  )
```

```{r}
#| label: fig-delivery-stages
#| fig-cap: "Share of the stage-0 delivery population retained at each DOB filter stage, by insurance category"

delivery_stage_effect |>
  select(year, ins, stage, n_mothers_pct_of_base, n_claims_pct_of_base) |>
  pivot_longer(
    cols = c(n_mothers_pct_of_base, n_claims_pct_of_base),
    names_to = "measure", values_to = "pct"
  ) |>
  mutate(measure = recode(measure,
    n_mothers_pct_of_base = "Mothers",
    n_claims_pct_of_base  = "Claims"
  )) |>
  ggplot(aes(x = year, y = pct, color = stage, linetype = measure)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ ins) +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = "% of stage 0", color = NULL, linetype = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom", legend.box = "vertical")
```

## Newborns: what each exclusion costs

```{r}
#| label: table-newborn-stages

newborn_stage_effect |>
  filter(ins == "Total") |>
  select(year, stage, n_newborns, n_newborns_drop, n_newborns_pct_of_base,
         n_claims, n_claims_drop, n_claims_pct_of_base) |>
  mutate(
    n_newborns_pct_of_base = percent(n_newborns_pct_of_base, accuracy = 0.1),
    n_claims_pct_of_base   = percent(n_claims_pct_of_base, accuracy = 0.1)
  ) |>
  gt(groupname_col = "year") |>
  tab_header(
    title    = "Newborns: Effect of DOB-Based Exclusions",
    subtitle = "All insurance categories combined; cumulative stages. Stage 0 = pre-DOB counts"
  ) |>
  cols_label(
    stage                  = "Stage",
    n_newborns             = "Newborns",
    n_newborns_drop        = "Newborns Lost",
    n_newborns_pct_of_base = "% of Stage 0",
    n_claims               = "Claims",
    n_claims_drop          = "Claims Lost",
    n_claims_pct_of_base   = "% of Stage 0"
  ) |>
  fmt_number(
    columns = c(n_newborns, n_newborns_drop, n_claims, n_claims_drop),
    decimals = 0
  )
```

```{r}
#| label: fig-newborn-stages
#| fig-cap: "Share of the stage-0 newborn population retained at each DOB filter stage, by insurance category"

newborn_stage_effect |>
  select(year, ins, stage, n_newborns_pct_of_base, n_claims_pct_of_base) |>
  pivot_longer(
    cols = c(n_newborns_pct_of_base, n_claims_pct_of_base),
    names_to = "measure", values_to = "pct"
  ) |>
  mutate(measure = recode(measure,
    n_newborns_pct_of_base = "Newborns",
    n_claims_pct_of_base   = "Claims"
  )) |>
  ggplot(aes(x = year, y = pct, color = stage, linetype = measure)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ ins) +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = "% of stage 0", color = NULL, linetype = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom", legend.box = "vertical")
```

## Claim counts across stages

Person counts and claim counts don't have to move together - an exclusion can
drop a small number of people carrying a large number of claims, or shave
out-of-window claims off people who stay in. This pulls the claim side out on
its own for both pipelines.

```{r}
#| label: claim-counts

claim_counts <- bind_rows(
  delivery_stage_effect |>
    filter(ins == "Total") |>
    transmute(
      pipeline = "Deliveries",
      year, stage_num, stage,
      n_claims, n_claims_drop, n_claims_pct_of_base,
      n_code_hits,
      n_persons = n_mothers
    ),
  newborn_stage_effect |>
    filter(ins == "Total") |>
    transmute(
      pipeline = "Newborns",
      year, stage_num, stage,
      n_claims, n_claims_drop, n_claims_pct_of_base,
      n_code_hits,
      n_persons = n_newborns
    )
) |>
  mutate(claims_per_person = n_claims / n_persons) |>
  arrange(pipeline, year, stage_num)

claim_counts
```

```{r}
#| label: table-claim-counts

claim_counts |>
  mutate(n_claims_pct_of_base = percent(n_claims_pct_of_base, accuracy = 0.1)) |>
  select(pipeline, year, stage, n_claims, n_claims_drop, n_claims_pct_of_base,
         n_code_hits, n_persons, claims_per_person) |>
  gt(groupname_col = "pipeline") |>
  tab_header(title = "Claim Counts by Filter Stage") |>
  cols_label(
    year                 = "Year",
    stage                = "Stage",
    n_claims             = "Claims",
    n_claims_drop        = "Claims Lost",
    n_claims_pct_of_base = "% of Stage 0",
    n_code_hits          = "Matched Code Rows",
    n_persons            = "Persons",
    claims_per_person    = "Claims / Person"
  ) |>
  fmt_number(columns = c(n_claims, n_claims_drop, n_code_hits, n_persons), decimals = 0) |>
  fmt_number(columns = claims_per_person, decimals = 2)
```

## Enrollment QC

With the enrollment age filters gone, a person with no `agg_yr_plan` row for
that year is no longer dropped - they land in `Not Enrolled` and then get folded
into `Other`. This sizes that bucket *before* the fold, so it's clear how much
of `Other` is genuinely unenrolled people rather than
Medicare/Federal coverage. A large or year-varying share here would undercut
the `Other` comparison against `self_pay + other + unknown_or_not_stated`.

```{r}
#| label: enrollment-qc

enrollment_qc <- bind_rows(
  delivery_stages_raw |>
    transmute(pipeline = "Deliveries", stage_num, stage, year, ins,
              n_persons = n_mothers, n_claims),
  newborn_stages_raw |>
    transmute(pipeline = "Newborns", stage_num, stage, year, ins,
              n_persons = n_newborns, n_claims)
) |>
  group_by(pipeline, stage_num, stage, year) |>
  # The 'Not Enrolled' subsets come FIRST: summarise() evaluates sequentially,
  # so if n_persons were reassigned above them they'd index the new scalar
  # rather than the original column.
  summarise(
    n_not_enrolled        = sum(n_persons[ins == "Not Enrolled"]),
    n_claims_not_enrolled = sum(n_claims[ins == "Not Enrolled"]),
    n_persons             = sum(n_persons),
    n_claims              = sum(n_claims),
    .groups = "drop"
  ) |>
  mutate(pct_not_enrolled = n_not_enrolled / n_persons) |>
  arrange(pipeline, year, stage_num)

enrollment_qc
```

```{r}
#| label: table-enrollment-qc

enrollment_qc |>
  mutate(pct_not_enrolled = percent(pct_not_enrolled, accuracy = 0.1)) |>
  select(pipeline, year, stage, n_persons, n_not_enrolled, pct_not_enrolled,
         n_claims, n_claims_not_enrolled) |>
  gt(groupname_col = "pipeline") |>
  tab_header(
    title    = "Unenrolled Share by Filter Stage",
    subtitle = "Persons with no agg_yr_plan match for that year, before folding into Other"
  ) |>
  cols_label(
    year                  = "Year",
    stage                 = "Stage",
    n_persons             = "Persons",
    n_not_enrolled        = "Not Enrolled",
    pct_not_enrolled      = "% Not Enrolled",
    n_claims              = "Claims",
    n_claims_not_enrolled = "Claims (Not Enrolled)"
  ) |>
  fmt_number(
    columns = c(n_persons, n_not_enrolled, n_claims, n_claims_not_enrolled),
    decimals = 0
  )
```

## DOB QC (non-cumulative)

Sizes each exclusion on its own terms: how many of the stage-0 population have
no DOB at all, how many have more than one, and - within each of those groups -
how many still have at least one qualifying claim (within 7 days of DOB for
infants, age 12-50 at DOS for mothers).

```{r}
#| label: pull-dob-qc

delivery_dob_qc <- dbGetQuery(uthealth, "
  SELECT year, ins, dob_status,
         n_mothers, n_mothers_age_ok, n_babies, n_claims, n_claims_age_ok
  FROM colour.delivery_dob_qc
  ORDER BY year, ins, dob_status
") |>
  as_tibble() |>
  rename(n_persons = n_mothers, n_persons_ok = n_mothers_age_ok,
         n_claims_ok = n_claims_age_ok) |>
  mutate(pipeline = "Deliveries") |>
  # n_babies is delivery-only; dropped here so the two pipelines stack cleanly
  select(-n_babies)

newborn_dob_qc <- dbGetQuery(uthealth, "
  SELECT year, ins, dob_status,
         n_newborns, n_newborns_in_window, n_claims, n_claims_in_window
  FROM colour.newborn_dob_qc
  ORDER BY year, ins, dob_status
") |>
  as_tibble() |>
  rename(n_persons = n_newborns, n_persons_ok = n_newborns_in_window,
         n_claims_ok = n_claims_in_window) |>
  mutate(pipeline = "Newborns")

dob_qc <- bind_rows(delivery_dob_qc, newborn_dob_qc) |>
  mutate(ins = if_else(ins == "Not Enrolled", "Other", ins)) |>
  group_by(pipeline, year, ins, dob_status) |>
  summarise(across(c(n_persons, n_persons_ok, n_claims, n_claims_ok), sum),
            .groups = "drop") |>
  group_by(pipeline, year) |>
  mutate(pct_of_year = n_persons / sum(n_persons)) |>
  ungroup() |>
  arrange(pipeline, year, ins, dob_status)

dob_qc
```

```{r}
#| label: table-dob-qc

dob_qc |>
  group_by(pipeline, year, dob_status) |>
  summarise(across(c(n_persons, n_persons_ok, n_claims, n_claims_ok), sum), .groups = "drop") |>
  group_by(pipeline, year) |>
  mutate(pct_of_year = n_persons / sum(n_persons)) |>
  ungroup() |>
  mutate(pct_of_year = percent(pct_of_year, accuracy = 0.1)) |>
  gt(groupname_col = "pipeline") |>
  tab_header(
    title    = "DOB Availability in the Stage-0 Population",
    subtitle = "All insurance categories combined"
  ) |>
  cols_label(
    year         = "Year",
    dob_status   = "DOB Status",
    n_persons    = "Persons",
    pct_of_year  = "% of Year",
    n_persons_ok = "Persons w/ Qualifying Claim",
    n_claims     = "Claims",
    n_claims_ok  = "Qualifying Claims"
  ) |>
  fmt_number(columns = c(n_persons, n_persons_ok, n_claims, n_claims_ok), decimals = 0)
```

## Filtered counts vs. vital statistics

The point of the exclusions is accuracy, not just attrition - this checks
whether each stage moves the claims counts toward or away from the vital
stats benchmark. The headline stage's rows here are the same numbers as the
Overall / Deliveries / Newborns tables above.

```{r}
#| label: stages-vs-vital

delivery_stages_vs_vital <- delivery_stages |>
  left_join(vital_long, by = c("year", "ins")) |>
  mutate(
    diff_babies     = n_babies - vital_n,
    pct_diff_babies = diff_babies / vital_n
  ) |>
  arrange(year, ins, stage_num)

newborn_stages_vs_vital <- newborn_stages |>
  left_join(vital_long, by = c("year", "ins")) |>
  mutate(
    diff_newborns     = n_newborns - vital_n,
    pct_diff_newborns = diff_newborns / vital_n
  ) |>
  arrange(year, ins, stage_num)

delivery_stages_vs_vital
newborn_stages_vs_vital
```

```{r}
#| label: fig-stages-vs-vital
#| fig-cap: "% difference from vital statistics births at each filter stage (Total across insurance categories)"

bind_rows(
  delivery_stages_vs_vital |>
    filter(ins == "Total") |>
    transmute(pipeline = "Deliveries (births)", year, stage, pct_diff = pct_diff_babies),
  newborn_stages_vs_vital |>
    filter(ins == "Total") |>
    transmute(pipeline = "Newborns", year, stage, pct_diff = pct_diff_newborns)
) |>
  ggplot(aes(x = year, y = pct_diff, color = stage)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_line() +
  geom_point() +
  facet_wrap(~ pipeline) +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = "% difference from vital stats", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Export to Excel

Ten tabs, each a real Excel Table (banded rows + filter dropdowns, via
`writeDataTable()`) except Steps, which is documentation. Tabs 1-3 are the
fully filtered counts; tabs 4-9 carry all four stages including the
unfiltered stage 0:

1. **Overall** - Total-row-only summary across both pipelines, filtered (first tab)
2. **Deliveries** - mother count + baby count vs. vital stats, by category + Total, filtered
3. **Newborns** - person count vs. vital stats, by category + Total, filtered
4. **Delivery Stages** - DOB filter waterfall for deliveries, stage 0 through 3
5. **Newborn Stages** - DOB filter waterfall for newborns, stage 0 through 3
6. **Claim Counts** - claim-side view of the waterfall, both pipelines
7. **DOB QC** - No DOB / Single DOB / Multiple DOB split of the stage-0 population
8. **Enrollment QC** - unenrolled share by stage, before folding into Other
9. **Stages vs Vital** - each stage's counts against the vital stats benchmark
10. **Steps** - methodology notes (last tab)

```{r}
#| label: export-to-excel

# -- EDIT THIS --------------------------------------------------
out_path <- here::here("claims_vs_vital_stats_comparison.xlsx")
# -----------------------------------------------------------------

wb <- createWorkbook()

table_style <- "TableStyleMedium9"

# Sheet name -> data frame. Order here is the tab order in the workbook.
sheets <- list(
  "Overall"          = overall_comparison,
  "Deliveries"       = delivery_comparison,
  "Newborns"         = newborn_comparison,
  "Delivery Stages"  = delivery_stage_effect,
  "Newborn Stages"   = newborn_stage_effect,
  "Claim Counts"     = claim_counts,
  "DOB QC"           = dob_qc,
  "Enrollment QC"    = enrollment_qc,
  "Stages vs Vital"  = bind_rows(
    delivery_stages_vs_vital |>
      transmute(pipeline = "Deliveries", stage_num, stage, year, ins,
                n = n_babies, vital_n, diff = diff_babies, pct_diff = pct_diff_babies),
    newborn_stages_vs_vital |>
      transmute(pipeline = "Newborns", stage_num, stage, year, ins,
                n = n_newborns, vital_n, diff = diff_newborns, pct_diff = pct_diff_newborns)
  )
)

# openxlsx's widths = "auto" is unreliable about leaving room for the
# header text itself (it can underestimate and clip column names), so
# size each column explicitly off the wider of the header name or its
# longest formatted value instead.
set_widths_for_headers <- function(wb, sheet, df) {
  if (nrow(df) == 0) return(invisible(NULL))
  col_widths <- vapply(seq_along(df), function(i) {
    header_len <- nchar(names(df)[i])
    data_len   <- suppressWarnings(max(nchar(format(df[[i]])), na.rm = TRUE))
    if (!is.finite(data_len)) data_len <- 0
    max(header_len, data_len) + 2
  }, numeric(1))
  setColWidths(wb, sheet, cols = seq_along(df), widths = col_widths)
}

# writeDataTable doesn't take a numFmt arg, so percent formatting is
# applied as a separate style pass on top of each table. Anything named
# pct_* or *_pct_of_base is a proportion and gets the 0.0% format.
pct_style <- createStyle(numFmt = "0.0%")

for (sheet_name in names(sheets)) {
  df <- sheets[[sheet_name]]

  addWorksheet(wb, sheet_name)
  writeDataTable(wb, sheet_name, df, tableStyle = table_style)

  pct_cols <- which(grepl("^pct_|_pct_of_base$|^pct_of_year$", names(df)))
  if (length(pct_cols) > 0 && nrow(df) > 0) {
    addStyle(
      wb, sheet_name, pct_style,
      rows = 2:(nrow(df) + 1),
      cols = pct_cols,
      gridExpand = TRUE
    )
  }

  set_widths_for_headers(wb, sheet_name, df)
}

# A note row above each filtered tab would break the Excel Table, so the
# headline stage is recorded on the Steps tab instead (see below).

# ------------------------------------------------------------------
# Steps tab (methodology notes - plain text, not a data table)
# ------------------------------------------------------------------

steps_text <- c(
  "Get counts and compare to vital stats",
  "",
  glue("Overall / Deliveries / Newborns tabs report stage {headline_stage}: {headline_label}"),
  "The Delivery Stages, Newborn Stages, Claim Counts and Stages vs Vital tabs carry",
  "every stage, including the unfiltered stage 0 (pre-DOB counts).",
  "",
  "Infants:",
  "- Get all claims using \"Newborn Codes\" tab of Orr codelists (CPT and DX)",
  "- No enrollment age filter (the old age < 1 filter has been removed)",
  "- Left join colour.dob on pers_id; keep dob_derv and the multiple-DOB flag",
  "- Qualifying claim = date of service within +/- 7 days of dob_derv",
  "- Get count distinct people",
  "",
  "Deliveries:",
  "- Get all claims using \"Delivery Codes\" tab of Orr codelists (CPT and DX)",
  "- No enrollment age filter (the old age > 12 filter has been removed)",
  "- Left join to enrollment for insurance; no match for the year = Not Enrolled",
  "- Left join colour.dob on pers_id; keep dob_derv and the multiple-DOB flag",
  "- Qualifying claim = age at date of service between 12 and 50 inclusive",
  "- Get one instance per person per year taking the maximum baby count for that year (for non-babycount codes, coalesce value of 1)",
  "",
  "DOB filter stages (cumulative, both pipelines):",
  "- Stage 0: all matched claims at any age - looser than the old pipeline,",
  "  which also applied an enrollment age filter. Stages 2-3 are comparable.",
  "- Stage 1: + person has a DOB in colour.dob",
  "- Stage 2: + person has at least one qualifying claim (7-day window / age 12-50)",
  "- Stage 3: + person has a single DOB only (colour.dob.mult_dob is null)",
  "- From stage 2 on, claim counts and baby counts use qualifying claims only;",
  "  n_babies_asis carries the unrestricted baby_count through every stage",
  "",
  "Counts reported at every stage:",
  "- Persons (count distinct pers_id)",
  "- Claims (count distinct clm_id) and matched code rows (a claim can hit several codes)",
  "- Births (sum of baby_count) for deliveries, both stage-restricted and as is",
  "",
  "Notes:",
  "- There is overlap of codes between deliveries and codesets",
  "- Age is now controlled only by the DOB criteria at stage 2, so stage 0 and 1",
  "  counts include people of any age with a matched code",
  "- Not Enrolled (no agg_yr_plan match for the year) is folded into Other to",
  "  match the 3-way vital stats split; see the Enrollment QC tab for its size",
  "- Vital Stats are recorded as the mother's insurance (not necessarily the baby's)",
  "- colour.dob is assumed unique on pers_id; the SQL collapses it defensively anyway",
  "- colour.dob.mult_dob is a null/1 flag: 1 = more than one DOB on file,",
  "  null = single DOB. The dobs array is carried as n_dobs for QC only",
  "- NY Vital Stats: Link",
  "- TX Vital Stats:"
)
header_rows <- which(steps_text %in% c(
  "Get counts and compare to vital stats",
  "Infants:",
  "Deliveries:",
  "DOB filter stages (cumulative, both pipelines):",
  "Counts reported at every stage:",
  "Notes:"
))
tx_row <- length(steps_text)

# -- EDIT THIS: the jsessionid pasted in chat is a temporary session
# token that will have expired - swap in a fresh/permanent query URL
# if you have one -----------------------------------------------
tx_url <- "https://wonder.cdc.gov/controller/datarequest/D149"
# -----------------------------------------------------------------

addWorksheet(wb, "Steps")
writeData(wb, "Steps", steps_text, colNames = FALSE)
addStyle(wb, "Steps", createStyle(textDecoration = "bold", fontSize = 12), rows = header_rows, cols = 1)

class(tx_url) <- "hyperlink"
writeData(wb, "Steps", x = tx_url, startCol = 2, startRow = tx_row)

setColWidths(wb, "Steps", cols = 1, widths = 100)
setColWidths(wb, "Steps", cols = 2, widths = 55)

saveWorkbook(wb, out_path, overwrite = TRUE)

cat("Wrote comparison workbook to", out_path, "\n")
```

```{r}
#| label: disconnect

dbDisconnect(uthealth)
```
